---
title: "Taxi Project Ver 2"
author: "Avery John"
date: "2026-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Import Libraries
```{r}
library(arrow)
library(dplyr)
library(lubridate)
library(ggplot2)
library(scales)
library(car)
```

#Import Parquet File
```{r}
#We decided to subset the data by focusing on 10% of the observations of a specific month. In this case, we're subsetting 10% of the data in January 2022.

taxi <- read_parquet(file.choose())
taxi %>%
  count() %>%
  collect() #2463931 observations

#Taking about 10% of the data as our sample
set.seed(123)
taxi_sample <- taxi %>%
  slice_sample(n = 250000) %>%
  collect()

taxi_sample %>%
  count() %>%
  collect()
```

#Exploring the Data
```{r}
str(taxi_sample)
summary(taxi_sample)
colSums(is.na(taxi_sample)) #Some NA values present in RatecodeID
```

#Cleaning the Data
```{r}
summary(taxi_sample$trip_distance)
#The max distance is 306,159 miles, which seems wildly off.

taxi_sample %>% filter(trip_distance == max(trip_distance))
#This observation had a total amount of $22.2, which doesn't correlate with the trip distance (unless they were traveling to the moon). Before further analysis, I decided to remove this extreme outlier. 

#Cleaning trip distance
taxi_clean <- taxi_sample %>%
  filter(trip_distance > 0,
         trip_distance < 100)

#Cleaning fare and tip amount
taxi_clean <- taxi_clean %>%
  filter(fare_amount > 0)

taxi_clean <- taxi_clean %>%
  filter(tip_amount >= 0)

#I created a trip duration variable to merge the dropoff and pickup time and convert them to minutes (correlates with trip distance)
taxi_clean <- taxi_clean %>%
  mutate(trip_duration = as.numeric(difftime(tpep_dropoff_datetime,
                                             tpep_pickup_datetime,
                                             units = "mins")))
str(taxi_clean) #New column called trip_duration
taxi_clean <- taxi_clean %>%
  filter(trip_duration > 0,
         trip_duration < 300) #Cap at 5 hours

#Cleaning RatecodeID (dropping null/99)
taxi_clean %>% count(RatecodeID)
taxi_clean <- taxi_clean %>%
  filter(!is.na(RatecodeID),
         RatecodeID != 99) 

#I also created a pickup hour variable to collect the pickup hour of each observation
taxi_clean$pickup_hour <- hour(taxi_clean$tpep_pickup_datetime)

#Then, I cleaned the payment type variable
taxi_clean <- taxi_clean %>%
  filter(payment_type == 1 | payment_type == 2)

str(taxi_clean) #New column called pickup_hour and payment_type filtered to credit card and cash
```

#Insight 1: Tipping behavior varies significantly by pickup zone, with zone 132 (JFK Airport) having a high average tip amount
```{r}
taxi_clean %>% count(PULocationID, sort = TRUE)
#Top 5 PU Zones: 236 (Manhattan UES North), 237 (Manhattan UES South), 132 (JFK Airport), 161 (Midtown Center), 186 (Penn Station)

top_5_zones <- taxi_clean %>% 
  filter(PULocationID == 132 | PULocationID == 161 | PULocationID == 186 | PULocationID == 236 | PULocationID == 237) %>%
  mutate(PULocationID = dplyr::recode(PULocationID,
                               '132' = 'JFK Airport',
                               '161' = 'Midtown Center',
                               '186' = 'Penn Station',
                               '236' = 'UES North',
                               '237' = 'UES South'))

summary_zones <- top_5_zones %>%
  group_by(PULocationID) %>%
  summarise(average_tip = mean(tip_amount, na.rm = TRUE))

ggplot(summary_zones, aes(x = factor(PULocationID), y = average_tip, fill = factor(PULocationID))) +
  stat_summary(fun = mean, geom = "bar") +
  geom_text(aes(label = round(average_tip, 2)), vjust = -0.5) +
  labs(title = "Average Tip Amount in Top 5 Pickup Zones",
       x = "Pickup Location",
       y = "Average Tip Amount") +
  theme_minimal() +
  theme(legend.position = "none")
```

#Insight 2: People who pay with credit card are more likely to tip than people who pay with cash
```{r}
taxi_clean %>% count(payment_type, sort = TRUE)

taxi_clean %>%
  mutate(payment_type = dplyr::recode(payment_type,
                                      '1' = 'Credit Card',
                                      '2' = 'Cash')) %>%
  group_by(payment_type) %>%
  summarise(avg_tip = mean(tip_amount, na.rm = TRUE)) %>%
  ggplot(aes(x = payment_type, y = avg_tip, fill = payment_type)) +
  geom_col() +
  geom_text(aes(label = round(avg_tip, 2)), vjust = -0.5) +
  scale_y_continuous(labels = scales::label_dollar()) +
  labs(title = "Average Tip Amount by Payment Type",
       x = "Payment Type",
       y = "Average Tip ($)") +
  theme(legend.position = "none")
```

#Insight 3: Higher fare amount and longer trip duration are both positively associated with higher average tip amount
```{r}
top_5_zones %>%
  mutate(fare_bin = cut(fare_amount,
                        breaks = c(0, 5, 20, 50, 100, 400))) %>%
  group_by(fare_bin) %>%
  summarise(avg_tip = mean(tip_amount, na.rm = TRUE)) %>%
  ggplot(aes(x = fare_bin, y = avg_tip, fill = fare_bin)) +
  geom_col() +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Average Tip Amount by Fare",
       x = "Fare Amount Levels ($)",
       y = "Average Tip ($)") +
  theme(legend.position = "none")

top_5_zones %>%
  mutate(duration_bin = cut(trip_duration,
                        breaks = c(0, 10, 15, 20, 100, 300))) %>%
  group_by(duration_bin) %>%
  summarise(avg_tip = mean(tip_amount, na.rm = TRUE)) %>%
  ggplot(aes(x = duration_bin, y = avg_tip, fill = duration_bin)) +
  geom_col() +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Average Tip Amount by Trip Duration",
       x = "Length of Trip (Minutes)",
       y = "Average Tip ($)") +
  theme(legend.position = "none")

#Fare amount and trip duration are dependent on one another. Longer trips typically result in higher fare amounts. 
```

#Insight 4: Tips are generally higher between 4pm and 7pm across the top 5 pickup zones
```{r}
top_5_zones$pickup_hour <- hour(top_5_zones$tpep_pickup_datetime)

hour_labels <- format(strptime(0:23, "%H"), "%I %p")
hour_labels <- sub("^0", "", hour_labels)

top_5_zones %>%
  group_by(pickup_hour) %>%
  summarise(avg_tip = mean(tip_amount, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(pickup_hour,
               levels = 0:23,
               labels = hour_labels), y = avg_tip, fill = "steelblue")) +
  geom_col() +
  scale_y_continuous(labels = label_dollar()) +
  labs(title = "Average Tip Amount by Hour",
       x = "Hour",
       y = "Average Tip ($)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(legend.position = "none")

```

#Multiple Linear Regression Model
```{r}
taxi_clean$PULocationID <- factor(taxi_clean$PULocationID)
taxi_clean$payment_type <- factor(taxi_clean$payment_type)

#Tipping data is often very skewed and not in line with the assumptions of linear regression. To get a model that satisfies the linear regression assumptions, the next step to take is to perform a log transformation.
taxi_clean <- taxi_clean %>%
  mutate(
    log_fare = log(fare_amount + 1),
    log_tip = log(tip_amount + 1),
    log_distance = log(trip_distance + 1),
    log_duration = log(trip_duration + 1)
  )

taxi_model_log <- lm(log_tip ~ log_fare + log_duration + payment_type + PULocationID + pickup_hour, data = taxi_clean)
summary(taxi_model_log)

#Error Metrics
pred_log <- predict(taxi_model_log, newdata = taxi_clean)
pred <- exp(pred_log) - 1
mae <- mean(abs(taxi_clean$tip_amount - pred))
print(mae)
rmse <- sqrt(mean((taxi_clean$tip_amount - pred)^2))
print(rmse)
vif(taxi_model_log, type = "predictor")
AIC(taxi_model_log)
plot(taxi_model_log)

#The residuals plot has a set of residuals that decrease in a linear fashion as the fitted values get bigger. To investigate this further, I took a small random sample of 200 observations from my current sample of around 245k to get a closer look at the specific observation points.

set.seed(123)
taxi_clean_small <- taxi_clean %>%
  slice_sample(n = 200) %>%
  collect()

taxi_model_2 <- lm(log_tip ~ log_fare + log_duration + payment_type + PULocationID + pickup_hour, data = taxi_clean_small)
summary(taxi_model_2)

plot(taxi_model_log_2)

ggplot(taxi_clean_small, aes(x = fitted(taxi_model_log_2), y = residuals(taxi_model_2))) +
  geom_point() +
  geom_text_repel(aes(label = rownames(taxi_clean_small)))


#The observations from the plot
taxi_clean_small[c(3, 32, 37, 55, 124, 134, 143, 150, 187), "PULocationID"]

#After taking a small subset of my sample, I see that the log tip for these observations all equal 0. This data has a mass constant of tips around $1, and the log of 1 is equal to 0, hence why it creates a downward residual line.

#Solution: Create a binary indicator variable for flat tipping behavior
taxi_clean$flat_tip <- ifelse(taxi_clean$tip_amount <= 1, 1, 0)
str(taxi_clean)

taxi_model_3 <- lm(log_tip ~ log_fare + log_duration + payment_type + pickup_hour + flat_tip, data = taxi_clean)
summary(final_model)

pred_log_3 <- predict(taxi_model_3, newdata = taxi_clean)
pred <- exp(pred_log_3) - 1
mae <- mean(abs(taxi_clean$tip_amount - pred))
print(mae)
rmse <- sqrt(mean((taxi_clean$tip_amount - pred)^2))
print(rmse)
vif(taxi_model_3, type = "predictor")
AIC(taxi_model_3)
plot(taxi_model_3)


#Still see the straight downward line on the residuals plot
#Plot Cook's Distance and remove high leverage outliers

cooks <- cooks.distance(taxi_model_3)
taxi_clean2 <- taxi_clean[cooks < (4 / nrow(taxi_clean)), ]

taxi_model_4 <- lm(log_tip ~ log_fare + log_duration + payment_type + pickup_hour + flat_tip, data = taxi_clean2)
summary(taxi_model_4)
pred_log_4 <- predict(taxi_model_4, newdata = taxi_clean2)
pred <- exp(pred_log_4) - 1
mae <- mean(abs(taxi_clean2$tip_amount - pred))
print(mae)
rmse <- sqrt(mean((taxi_clean2$tip_amount - pred)^2))
print(rmse)
vif(taxi_model_3, type = "predictor")
AIC(taxi_model_3)

plot(taxi_model_4)
```